Storm surge modelling in the Strait of Georgia

### 2014-12-05 RoughWork/manuscript.tex
final search for typos

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too high in the Strait of Juan de Fuca and too low in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

@@

Freshwater runoff from each watershed was divided between the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0-3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface air temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

@@

In general, there is a very good agreement between the observed and modelled residuals \ref{fig:feb2006}. The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual is in sync with observations. The errors in maximum residual range from 6.0 cm too high at Victoria to 15.9 cm too low at Campbell River. Both Point Atkinson and Patricia Bay experience a very good match with observed residuals with errors in the maximum less than 2 cm.

@@

Next, an assessment of the factors that are most important in scenarios leading up to storm surges is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}).Four simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly. Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the sea surface height anomaly. Fourth, a simulation with only wind forcing, that is no anomaly and no inverse barometer effect due to pressure. It is clear that the sea surface height anomaly contributes most significantly to the surge at each of these locations (Figure \ref{fig:factors}). When the sea surface height anomaly is not included as a forcing condition, the surge drops from a maximum 82 cm to 11 cm at Point Atkinson and its maximum occurs eight hours later.  A similar drop in the maximum surge and delay in timing of the maximum surge is observed at the other locations. The atmospheric model indicates southeasterly winds near Point Atkinson during the onset of the surge but after the peak surge has passed the winds shifted to westerly. A strong westerly wind would cause an elevated sea surface height at Point Atkinson as water piles against the coast, explaining the delay in peak surge when only atmospheric forcing is used since the winds shift to westerly after the peak surge has passed. A simulation with only wind forcing produces a surge of 7.8 cm at Point Atkinson, a small fraction of the total observed surge, and only centimetres at the other locations. Note that this wind-induced surge occurs well after the peak surge occurred in both the observations and simulations with full forcing.

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included, in contrast to other findings that reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}.

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. In a model domain that does not include the Pacific Ocean, this effect must be included as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the inverse barometer effect and the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain.

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44 cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. All Willmott skill scores rank above 0.89 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

The modelled water levels and residuals are too high for this hindcast, except at Victoria (Table \ref{tab:statistics}). Although Victoria has the smallest mean error in both the water level and residuals, its Willmott skill score and $\gamma^2$ values rank low when compared to the other stations. Generally, the skills scores are very good, scoring above 0.89 at all stations.

@@

Both of the simulations agree well with the observations (\ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

>>>>

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

@@

Freshwater runoff from each watershed was divided between the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface air temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

@@

In general, there is a very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual is in sync with observations. The errors in maximum residual range from 6.0 cm too high at Victoria to 15.9 cm too low at Campbell River. Both Point Atkinson and Patricia Bay experience a very good match with observed residuals with errors in the maximum less than 2 cm.

@@

Next, an assessment of the factors that are most important in scenarios leading up to storm surges is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Four simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly. Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the sea surface height anomaly. Fourth, a simulation with only wind forcing, that is no anomaly and no inverse barometer effect due to pressure. It is clear that the sea surface height anomaly contributes most significantly to the surge at each of these locations (Figure \ref{fig:factors}). When the sea surface height anomaly is not included as a forcing condition, the surge drops from a maximum 82 cm to 11 cm at Point Atkinson and its maximum occurs eight hours later.  A similar drop in the maximum surge and delay in timing of the maximum surge is observed at the other locations. The atmospheric model indicates southeasterly winds near Point Atkinson during the onset of the surge but after the peak surge has passed the winds shifted to westerly. A strong westerly wind would cause an elevated sea surface height at Point Atkinson as water piles against the coast, explaining the delay in peak surge when only atmospheric forcing is used since the winds shift to westerly after the peak surge has passed. A simulation with only wind forcing produces a surge of 7.8 cm at Point Atkinson, a small fraction of the total observed surge, and only centimetres at the other locations. Note that this wind-induced surge occurs well after the peak surge occurred in both the observations and simulations with full forcing.

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not include. Other findings have reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}, but only a minor effect is noticeable here.

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. In a model domain that does not include the Pacific Ocean, this effect must be included as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain.

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44~cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

The modelled water levels and residuals are too high for this hindcast, except at Victoria (Table \ref{tab:statistics}). Although Victoria has the smallest mean error in both the water level and residuals, its Willmott skill score and $\gamma^2$ values rank low when compared to the other stations. Generally, the skills scores are very good, scoring above 0.91 at all stations.

@@

Both of the simulations agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.
